---
title: "README"
author: "Pourya Naderi (Hide Lab)"
date: "9/16/2021"
bibliography: vignettes/references.bib  
output:
  md_document:
    variant: gfm
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<!-- badges: start -->
[![R-CMD-check](https://github.com/pouryany/PanomiR/workflows/R-CMD-check/badge.svg)](https://github.com/pouryany/PanomiR/actions)
[![lint](https://github.com/pouryany/PanomiR/workflows/lint/badge.svg)](https://github.com/pouryany/PanomiR/actions)
<!-- badges: end -->

## Introduction

PanomiR is a package for pathway and microRNA Analysis of gene expression data.
This document provides details about how to install and utilize various 
functionality in PanomiR.

For questions, comments, and other queries, contact pouryany@gmail.com

## Installation

PanomiR can be accessed via Bioconductor. To install, start R
and run the following code.

```{r installation_bioc, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("PanomiR")
```

You can also install the latest development version of PanomiR using GitHub.

```{r installation_git, eval = FALSE}
devtools::install_github("pouryany/PanomiR")
```
## Overview

PanomiR is a pipeline to prioritize disease-associated miRNAs based on activity 
of disease-associated pathways. The input datasets for PanomiR are (a) a gene 
expression disease dataset along with covariates, (b) a background collection of 
pathways/genesets, and (c) a collection of miRNAs containing gene targets.

The general workflow of PanomiR is (a) generation of pathway summary statistics 
from gene expression data, (b) detection of differentially activated pathways,
(c) finding coherent groups, or clusters, of differentially activated pathways,
and (d) detecting miRNAs targeting each group of pathways. 

Individual steps of the workflow can be used in isolation to carry out different
analyses. The following sections outline each step and material needed to
execute PanomiR. 

## 1. Pathway summarization

PanomiR can generate pathway activity profiles given a gene expression dataset
and a list of pathways.Pathway summaries are numbers that represent the overall activity of genes that 
belong to each pathway. These numbers are calculated based on a methodology
previously described in Altschuler et al. 2013 and
Joachim and Altschuler et al. 2018 [-@altschuler2013pathprinting; -@joachim2018relative].
Briefly, genes in each sample are ranked by their expression values and then
pathway summaries are calculated as the average rank-squared of genes within a 
pathway. The summaries are then center and scaled (zNormalized) across samples.

The default list of background pathways in PanomiR is formatted into a table
(`data("path_gene_table")`). The table is based on canonical pathways collection
of Molecular Signatures Database (MSigDB) V6.2 and it contains annotated 
pathways from a variety of sources excluding KEGG dataset [-@liberzon2011molecular].

** Users interested in using other pathway/geneset backgrounds, such as newer 
versions of MSigDB or KEGG, should refer to the [appendix](#geneset) of this manual.

This section uses a reduced example dataset from The Cancer Genome Atlas (TCGA)
Liver Hepatocellular Carcinoma (LIHC) dataset to generate
Pathway summary statistics [-@ally2017comprehensive]. 

```{r load_package}
library(PanomiR)

# Pathway reference from the PanomiR package
data("path_gene_table")
data("miniTestsPanomiR")
# Generating pathway summary statistics 

summaries <- pathwaySummary(miniTestsPanomiR$mini_LIHC_Exp,
                            path_gene_table, method = "x2",
                            zNormalize = TRUE)

head(summaries)[,1:2]
```




## 2. Differential Pathway activation

Once you generate the pathway activity profiles, as discussed in the last
section, there are several analysis that you can perform. We have bundled some 
of the most important ones into standalone functions. Here, we describe
differential pathway activation profiling, which is examining differences in
pathway activity profiles in user-determined conditions.

At this stage you need to provide a pathway-gene association table, an
expression dataset, and a covariates table. You need to specity what covariates
you would like to contrast. You also need to provide a contrast, as formatted in
limma. If the contrast is not provided, the function assumes the first two 
levels of the provided contrast covariate. **Note:** make sure the contrast 
covariate is formatted as factor.


```{r differential}

output0 <- differentialPathwayAnalysis(
                        geneCounts = miniTestsPanomiR$mini_LIHC_Exp,
                        pathways =  path_gene_table,
                        covariates = miniTestsPanomiR$mini_LIHC_Cov,
                        condition = 'shortLetterCode')

de.paths <- output0$DEP

head(de.paths)
```

## 3. Finding clusters of pathways

PanomiR provides a function to find groups coordinated differentially 
activated pathways based on a pathway co-expression network (PCxN) previously
described in Pita-Juárez and Altschuler et al [-@pita2018pathway]. Briefly, PCxN
is a network where nodes are pathways and links are co-expression between the
nodes. It is formatted into a table were rows represent edges. The edges of PCxN
are marked by two numbers, 1- a correlation co-efficient and 2- a significance
adjusted p-value. Cut-offs for both of these numbers can be manually set using 
PanomiR functions. See function manuals for more info. 

PCxN and its associated genesets are already released and can be accessed through
[pcxn Bioconductor package](http://bioconductor.org/packages/release/bioc/html/pcxn.html) and [pcxnData package](http://bioconductor.org/packages/release/data/experiment/html/pcxnData.html).

Here we have provided a small version of PCxN for tutorial purposes. A more 
recent version of PCxN based on MSigDB V6.2 is available through the 
data repository accompanying PanomiR manuscript, which can be found [here](https://github.com/pouryany/PanomiR_paper).


```{r pcxn}


# using an updated version of pcxn 

set.seed(2)
pathwayClustsLIHC <- mappingPathwaysClusters(
                            pcxn = miniTestsPanomiR$miniPCXN, 
                            dePathways = de.paths[1:300,],
                            topPathways = 200,
                            outDir=".",
                            plot = FALSE,
                            subplot = FALSE,
                            prefix='',
                            clusteringFunction = "cluster_louvain",
                            correlationCutOff = 0.1)


head(pathwayClustsLIHC$Clustering)

```


## 4. Prioritizing miRNAs per cluster of pathways.

PanomiR identifies miRNAs that target clusters of pathways, as defined in the 
last section. In order to this, you would need a reference table of
miRNA-Pathway association score (enrichment). We recommend using a customized
miRNA-Pathway association table, tailored to your experimental data. Here, 
we provide a pre-processed table for LIHC table and in the next section, we will
explain how to generate the customized tables.

Note that in the example below, we use a sampling rate of 50, the recommended
rate is between 500-1000. Also, we set the saveSampling argument to FALSE. This
argument should be set to TRUE if you wish to save your sampling and check for
different outputs from the clustering algorithms or pathway thresholds.


```{r miRNA}



set.seed(1)
output2 <- prioritizeMicroRNA(enriches0 = miniTestsPanomiR$miniEnrich,
                    pathClust = miniTestsPanomiR$miniPathClusts$Clustering,
                    topClust = 1,
                    sampRate = 50, 
                    method = c("aggInv"),
                    outDir = "Output/",
                    dataDir = "outData/",
                    saveSampling = FALSE,
                    runJackKnife = FALSE,
                    numCores = 1,
                    prefix = "outmiR",
                    saveCSV = FALSE)

head(output2$Cluster1)

```



## 5. miRNA-Pathway enrichment tables {#geneset}

PanomiR best performs on tissue/experiment-customized datasets. In order to do
this, you need to create a customized enrichment table. You can simply do so by
using the pathway and miRNA list that we have provided as a part of the package.
simply, plug in the name of the genes present (expressed) in your experiment in
the following code



```{r customized_mir, eval=FALSE}


# using an updated version of pcxn 
data("msigdb_c2")
data("targetScan_03")


customeTableEnrich <- miRNAPathwayEnrichment(mirSets = targetScan_03,
                                              pathwaySets = msigdb_c2,
                                              geneSelection = yourGenes,
                                              mirSelection = yourMicroRNAs,
                                              fromID = "ENSEMBL",
                                              toID = "ENTREZID",
                                              minPathSize = 9,
                                              numCores = 1,
                                              outDir = ".",
                                              saveOutName = NULL)

```

In the above section, the field ```fromID``` denotes the gene representation 
format of your input list. Here is a quick example that runs fast.
```{r customized_mir2}


# using an updated version of pcxn 
data("msigdb_c2")
data("targetScan_03")

tempEnrich <-miRNAPathwayEnrichment(targetScan_03[1:30],msigdb_c2[1:30])

head(tempEnrich)
```

## Session info
```{r sessionInfo}
sessionInfo()

```